A correct by construction hierarchy of reflective tactics for proving equalities in algebraic structures

### 2014-07-07 newPaper/paper.tex
paper continued

\begin{code}[caption=Production of the left associative form, captionpos=b, label=lst1:haskell2]
expr_l : Expr G x 
         -> (x' ** (LExpr G x', x = x'))
expr_l Zero = (_ ** (LZero, refl))
expr_l (Var i) = (_ ** (LPlus LZero i, refl))
expr_l (Plus ex ey) = 
  let (xl ** (xr, xprf)) = expr_l ex in
  let (yl ** (yr, yprf)) = expr_l ey in
    plusLExpr _ _ xr yr xprf yprf
      where 
      plusLExpr : (x', y' : Nat)
            -> {G : Vect n Nat} -> {x, y : Nat} 
            -> LExpr G x -> LExpr G y 
            -> (x' = x) -> (y' = y) 
            -> (w' ** (LExpr G w', x'+y'=w'))
      plusLExpr x' y' rx (LPlus e i) xprf yprf =
        let (x ** (rec, prf)) = 
          plusLExpr _ _ rx e refl refl in
          (_ ** (LPlus rec i, ?plusLExpr1))
      plusLExpr x' y' rx LZero xprf yprf =
        (_ ** (rx, ?plusLExpr2))

@@

We notice that for transforming the expression into its left associative equivalent representation, we've effectively needed to know where the variables and the $Z$ constants are : the functions $expr\_l$ and $l\_expr$ are doing different treatments for these different possibilities. \\
\\
We've got two metavariables to prove. The metavariable $plusLExpr1$ requires us to prove the goal : $x' + y' = x2 + index\ i\ G$ in a context where we've got, amongst other things,  $(xprf\ :\ x'\ =\ x)$, $(yprf\ :\ y'\ =\ x1\ +\ index\ i\ G)$ and $(prf\ :\ x\ +\ x1\ =\ x2)$.
Proving this goal uses the property of associativity after rewriting the goal with these three proof of equality $xprf$, $yprf$ and $prf$.

@@

And if we ask for the evaluation of this term, we should obtain $Just$ and a proof of equality between the two underlying numbers.
\begin{code}[caption=Evaluation of the result, captionpos=b, label=lst1:haskell2]
\x => \y => \z => e1_e2_testEq x y z

\x => \y => \z => Just (replace (sym (replace 
(sym (replace (replace (plusZeroRightNeutral x)
 refl (replace (sym (plusAssociative x [] y)) 
 refl))) (replace (sym (replace (replace 
 (plusZeroRightNeutral [...] refl)))) refl)) 
: (x : List Int) -> (y : List Int) -> 
  (z : List Int) -> 
  Maybe ((x + y) + x + z 
         = x+ (y + x) + z)
\end{code}

And we effectively get the proof of equality we wanted. As expected, this proof uses the properties of associativity ($plusAssociative$) and the property of neutrality of $Z$ for $plus$ ($plusZeroRightAssociative$).

@@

Now that we have introduced our ideas on this little example, it is time to apply them for our main goal : the implementation of a hierarchy of tactics proving equalities in algebraic structures. Very often, the properties available on a given type are the one of a well known structure : magma, semi-group, monoid, group, commutative-group...  We will no longer only work with the properties of associativity and neutral elements for Lists as we did in the previous section, but we will have available the properties of different structures, and these tactics will usable for any type that satisfies these properties.
In a Magma, we only have a $Plus$ operation, and no axioms with it. With a Semi-Group, we will have one axiom about this Plus : the property of associativity. In a Monoid, we will have one more property : the fact that we have a neutral element. In fact, this property will be expressed as 3 facts : there is a value $Zero$, and we've got $\forall x, Zero + x = x$ and $\forall x, x + Zero = x$. \\

@@

Note that this equality is a "week" equality : it only produces a proof when the two elements are equal, but it doesn't produce a proof of dis-equality when they are different. That's quite natural, since we want to generate proof of equality, and not to generate counter examples for dis-equalities, which is another problem.

@@

A Group is a Monoid, with two new operations. The binary operation $Minus$, and the unary operation $Neg$. One axiom says that the $Minus$ can always be simplified with the $Neg$ and the $Plus$. That means that $Minus$ is not a "primitive" operation of a Group, since we can always rewrite $(a-b)$ into $(a\ +\ -b)$.
The second axiom, which is the most important one for a Group, is the fact that any $c1$ admits $Neg\ c1$ as an inverse, where being an inverse if the following property :

\begin{code}[caption=Group, captionpos=b, label=lst1:haskell2]   
-- This is just a conjunctive predicate
hasSymmetric : (c:Type) -> (p:dataTypes.Monoid c) 
                -> c -> c -> Type
hasSymmetric c p a b = 
         (Plus a b = Zero, Plus b a = Zero)

class dataTypes.Monoid c => dataTypes.Group c where
	Minus : c -> c -> c
	Neg : c -> c
	Minus_simpl : (c1:c) -> (c2:c) -> 
	             Minus c1 c2 = Plus c1 (Neg c2) 
	Plus_inverse : (c1:c) -> 
	             hasSymmetric c (instance) c1 (Neg c1)
\end{code}

@@

As discussed in the section above, the algorithm of normalization will not directly use the concrete value of type $T$, but a reflected term, indexed over the concrete value.

>>>>

\begin{code}[caption=Production of the left associative form, captionpos=b, label=lst1:haskell2]
expr_l : Expr G x 
         -> (x' ** (LExpr G x', x = x'))
expr_l Zero = (_ ** (LZero, refl))
expr_l (Var i) = (_ ** (LPlus LZero i, refl))
expr_l (Plus ex ey) = 
  let (xl ** (xr, xprf)) = expr_l ex in
  let (yl ** (yr, yprf)) = expr_l ey in
    plusLExpr _ _ xr yr xprf yprf
      where 
      plusLExpr : (x', y' : Nat)
            -> {G : Vect n Nat} -> {x, y : Nat} 
            -> LExpr G x -> LExpr G y 
            -> (x' = x) -> (y' = y) 
            -> (w' ** (LExpr G w', x'+y'=w'))
      plusLExpr x' y' rx (LPlus e i) xprf yprf =
        let (xrec ** (rec, prf)) = 
          plusLExpr _ _ rx e refl refl in
          (_ ** (LPlus rec i, ?plusLExpr1))
      plusLExpr x' y' rx LZero xprf yprf =
        (_ ** (rx, ?plusLExpr2))

@@

We notice that for transforming the expression into its left associative equivalent representation, we've effectively needed to know where the variables and the $Z$ constants are : the functions $expr\_l$ and $l\_expr$ are doing different treatments for these different possibilities. \\
\\
We've got two metavariables to prove. The metavariable $plusLExpr1$ requires us to prove the goal : $x' + y' = xrec + index\ i\ G$ in a context where we've got, amongst other things,  $(xprf\ :\ x'\ =\ x)$, $(yprf\ :\ y'\ =\ x1\ +\ index\ i\ G)$ and $(prf\ :\ x\ +\ x1\ =\ xrec)$.
Proving this goal uses the property of associativity after rewriting the goal with these three proof of equality $xprf$, $yprf$ and $prf$.

@@

And if we ask for the evaluation of this term, we should obtain $Just$ and a proof of equality between the two underlying numbers.
\begin{code}[caption=Obtained proof, captionpos=b, label=lst1:haskell2]
#\x => \y => \z => e1_e2_testEq x y z

\x => \y => \z => Just (replace (sym (replace 
  (sym (replace (plusZeroRightNeutral x) 
  (plusAssociative x 0 y))) (replace 
  (sym (replace (plusZeroRightNeutral x) 
  (plusAssociative x 0 z))) (replace (replace 
  (plusZeroRightNeutral (plus xy)) 
  (plusAssociative (plus x y) 0 x)) [...]
: (x : Nat) -> (y : Nat) -> (z : Nat) 
  -> Maybe ((x + y) + x + z 
            = x+ (y + x) + z)
\end{code}

And we effectively get the proof of equality we wanted. As expected, this proof uses the properties of associativity ($plusAssociative$) and the property of neutrality of $Z$ for $plus$ ($plusZeroRightNeutral$).

@@

Now that we have introduced our ideas on this little example, it is time to apply them for our main goal, more general : the implementation of a hierarchy of tactics proving equalities in algebraic structures. Very often, the properties available on a given type are the one of a well known structure : magma, semi-group, monoid, group, commutative-group...  We will no longer only work with the properties of associativity and of right neutral element for natural numbers as we did in the previous section, but we will have available the properties of different structures, and these tactics will usable for any type that satisfies these properties.

@@

Note that this equality is a "week" equality : it only produces a proof when the two elements are equal, but it doesn't produce a proof of dis-equality when they are different -- instead, it simply produces the value $Nothing$--. That's quite natural, since we want to generate proof of equality, and not to generate counter examples for proving dis-equalities, which is another problem.

@@

A Group is a Monoid, with two new operations. The binary operation $Minus$, and the unary operation $Neg$. One axiom says that the $Minus$ can always be simplified with the $Neg$ and the $Plus$. That means that $Minus$ is not a "primitive" operation of a Group, since we can always rewrite $(a-b)$ into $(a\ +\ -b)$.
The second axiom, which is the most important one for a Group, is the fact that any value $c1$ admits $Neg\ c1$ for inverse, where being an inverse is the following property :

\begin{code}[caption=Group, captionpos=b, label=lst1:haskell2]   
-- This is just a conjunctive predicate
hasSymmetric : (c:Type) -> (p:Monoid c) 
                -> c -> c -> Type
hasSymmetric c p a b = 
         (Plus a b = Zero, Plus b a = Zero)

class Monoid c => Group c where
	Minus : c -> c -> c
	Neg : c -> c
	Minus_simpl : (c1:c) -> (c2:c) -> 
	             Minus c1 c2 = Plus c1 (Neg c2) 
	Plus_inverse : (c1:c) -> 
	             hasSymmetric c _ c1 (Neg c1)
\end{code}

@@

As discussed in the section 2.1 on the smaller problem, the algorithm of normalization will not directly use the concrete value of type $T$, but a reflected term, indexed over the concrete value.
